### 2016-03-22 paper/paper.tex
please be here
fiddling with paper, using google to run everything

\documentclass{article}
\usepackage{minted}
\usemintedstyle{emacs}

@@

% question: Can macros be used to create interesting and useful parallelism
% tools for recursive functions in clojure.
\section{Introduction}
% most important section in paper:
% - introduce the problem being solved
% - prove that this problem is interesting
% - why is this hard?
% - how does it differ from other approaches?
% - what are key results/limitations
Clojure is a lisp--like language running on the Java Virtual Machine.
The language's default immutable data structures and it's sophisticated STM system make it well suited for parallel and concurrent programming.
Additionally, because Clojure runs on the Java Virtual Machine, Clojure developers can take advantage of existing cross--platform parallelism libraries, such as Java's excellent ExecutorService framework, to write parallel code.

@@

\section{Macros}
\subsection{parlet}
The first of the parallel macros is called parlet.
Parlet is a modified version of the Clojure let form.
The parlet macro has exactly the same semitics as Clojure's let, but it evaluates all of
it's bindings in parallel (if there is no dependencies between them).
For example, suppose I had some long running function foo, if I needed to add
the result of two calls to this function, I could do the following to execute my long running calls in parallel:

% TODO use a figure
\begin{minted}{clojure}
    (parlet
      [a (foo value1)
       b (foo value2)]
       ... ; some other code here
      (+ a b))
\end{minted}

The execution of each of the function calls done in a ForkJoinPool.
The calls to foo both start in the background, then we attempt to evaluate the body of the let.
This means that the code in the body of the let which does not depend on the computations of a and b can execute without delay.
This is obviously only safe when foo does not have side effects which may impact the other code in the function.
In Clojure, it is very common (although not guaranteed by the compiler) for function calls to be pure.

Nested parlet macros will also behave quite well.
For example:

\begin{minted}{clojure}
    (parlet
      [a (foo 100)]
      ;; some other code not using a
      (parlet
        [b (foo 200)]
        (+ a b)))
\end{minted}

Both calls to the function foo will be processed in the background, until we reach the line using the variables that the results are bound to.
This means that parlets can be nested with little concern (again as long as functions are pure).

@@

The parlet macro also supports dependency detection.
Clojure let forms can use names defined previously in the let, and the bindings are evaluated from first to last.
The following form is valid (the binding to b used the value defined in the first binding):

\begin{minted}{clojure}
    (let
     [a 1
      b (+ 1 a)]
     a)
\end{minted}

This form would evaluate to 2.
If we plan on evaluating each binding in parallel, we can't allow the bindings to have dependencies.
So, the parlet macro looks through the bindings in the macro to determine if any of them depend on
any of the names defined previously in the macro.
If there are any, the macro will halt the compiler and report an error to the user.

@@

Parexpr breaks down expressions and (aggressively) evaluates them in parallel.
parexpr uses parlet to parallelize forms.
Suppose again that I had a long running function foo, I wanted to call it twice, and add the results.
With parexpr the following will make both calls to foo in parallel:

\begin{minted}{clojure}
    (parexp (+ (foo value1) (foo value2)))
\end{minted}

@@

>>>>

\documentclass{sig-alternate}
\usepackage{minted}
\usemintedstyle{emacs}

\def\mintcode#1{\mbox{\mintinline{clojure}{#1}}}
\def\code#1{\mbox{\texttt{#1}}}
\def\parlet{\code{parlet}}

@@

% question: Can macros be used to create interesting and useful parallelism
% tools for recursive functions in clojure.
\section{Introduction}
% most important section in paper:
% - introduce the problem being solved
% - prove that this problem is interesting
% - why is this hard?
% - how does it differ from other approaches?
% - what are key results/limitations
Clojure is a lisp--like language running on the Java Virtual Machine.
The language's default immutable data structures and it's sophisticated STM system make it well suited for parallel and concurrent programming.
Additionally, because Clojure runs on the Java Virtual Machine, Clojure developers can take advantage of existing cross--platform parallelism libraries, such as Java's excellent \code{ExecutorService} framework, to write parallel code.

@@

\section{Macros}
\subsection{parlet}
The first of the parallel macros is called \parlet{}.
\parlet{} is a modified version of the Clojure let form.
The \parlet{} macro has exactly the same semantics as Clojure's \code{let}, but it evaluates all of it's bindings in parallel.
For example, suppose I had some long running function \code{foo}.
I need to add the result of two calls to this function.
In Figure \ref{fig:parlet_basic}, we use \parlet{} to make two calls to \code{foo}, then add the results.

\begin{figure}[h]
    \begin{minted}{clojure}
        (parlet
          [a (foo value1)
           b (foo value2)]
           ... ; some other code here
          (+ a b))
    \end{minted}
    \caption{Example of a parlet form}
    \label{fig:parlet_basic}
\end{figure}

In this example, the expressions \mintcode{(foo value1)} and \mintcode{(foo value2)} are both evaluated as \code{ForkJoinTask}s in a \code{ForkJoinPool}.

The calls to \code{foo} are both \code{forked} immediately, then we attempt to evaluate the body of the \code{let}.
This means that the code in the body of the \code{let} which does not depend on the computations of \code{a} and \code{b} can execute without delay.

This transformation is obviously only safe when \code{foo} does not have side effects which may impact the other code in the function, or the values of terms used in subsequent let terms.
In Clojure, it is very common (although not guaranteed by the compiler) for function calls to be pure.

Nested parlet macros will also behave quite well.

\begin{figure}[h]
    \begin{minted}{clojure}
        (parlet
          [a (foo 100)]
          ;; some other code not using a
          (parlet
            [b (foo 200)]
            (+ a b)))
    \end{minted}
    \caption{Nested parlet forms}
    \label{fig:parlet_noblock}
\end{figure}

In Figure \ref{fig:parlet_noblock}, calls to the function foo will be processed in the background until we reach the line using the variables that the results are bound to.
This means that parlets can be nested with little concern (again as long as functions are pure).

@@

The parlet macro also supports dependency detection.
Clojure let forms can use names defined previously in the let, and the bindings are evaluated from first to last.

\begin{figure}[h]
    \begin{minted}{clojure}
        (parlet
        [a 1
         b (+ 1 a)]
         a)
    \end{minted}
    \caption{A parlet containing a dependency}
    \label{fig:parlet_dep}
\end{figure}

Without the parlet, the let form in Figure \ref{fig:parlet_dep} would evaluate to 2.
If we plan on evaluating each binding in parallel, we can't allow the bindings to have dependencies.
So, the parlet macro looks through the bindings in the macro to determine if any of them depend on
any of the names defined previously in the macro.
If there are any, the macro will halt the compiler and report an error to the user.

@@

Parexpr breaks down expressions and (aggressively) evaluates them in parallel.
parexpr uses parlet to parallelize forms.
Suppose again that I had a long running function foo, I wanted to call it twice, and add the results.
The code in Figure \ref{fig:parexpr} will make both of the long running calls to foo in parallel.

\begin{figure}[h]
    \begin{minted}{clojure}
        (parexp (+ (foo value1) (foo value2)))
    \end{minted}
    \caption{Example using parexpr to evaluate long running functions}
    \label{fig:parexpr}
\end{figure}

@@

\section{Benchmarking}
\subsection{Setup}

\section{Results}

\section{Conclusions}
% the jvm is (mostly) awesome
% immutability makes things like this really straightforward
